---
title: "Negative Ties"
subtitle: "The occurrence and mechanisms of positive and negative ties in the  social networks"
author: "Mohamed Shehata"
date: "02/04/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```




```{r,echo=FALSE,message=FALSE,warning=FALSE,results='hide'}

library(knitr)
library(sna)

#Prepare the data
#first upload the data as we can only use two networks for time 1 and time 4 as both have the behavior and attitude for all nodes in the network 

load("Final project Dataset.RData")

#Delete the whole nodes that actor non-response coded with = 9, as some students dropout in network 4 and network 5 

present4 <- rowSums(t4.dat) != (nrow(t4.dat) * 9)

t4.data  <- t4.dat[present4 ,present4 ] 

present1 <- rowSums(t1.dat) != (nrow(t1.dat) * 9)

t1.data  <- t1.dat[present1 ,present1 ] 

present5 <- rowSums(t5.dat) != (nrow(t5.dat) * 9)

t5.data  <- t5.dat[present5 ,present5 ] 

#create out matrixes and recoded it again as we will have a positive and negative binary networks for both the 4th and the 5th network

#the positive networks will be coded with 1= for friends and best friends and friendly relation (1 &2 &3) and otherwise 0 

#network 4 
Positive_data4 <- ifelse(t4.data == 1, yes = 1, 
                         ifelse(t4.data == 2, yes = 1,
                         ifelse(t4.data == 3, yes = 1,
                                no = 0)))

#network 5

Positive_data5 <- ifelse(t5.data == 1, yes = 1, 
                         ifelse(t5.data == 2, yes = 1,
                         ifelse(t5.data == 3, yes = 1,
                                no = 0)))

#then, we will also recode for the negative networks as will be coded with 1 for troubled relation = 5 and otherwise 0 

negative_data4 <- ifelse(t4.data == 5, 1, 0)

#network 5


negative_data5 <- ifelse(t5.data == 5, 1, 0)


#Create two networks from these data frames that we are interested in 

network1 <- as.network(t1.data)

network4 <- as.network(t4.data)

network5 <- as.network(t5.data)

network4_positive <- as.network(Positive_data4)

network4_negative <- as.network(negative_data4)
  
network5_positive <- as.network(Positive_data5)

network5_negative <- as.network(negative_data5)


#create our attribute variables from cov1.dat to avector to be used later with #networks if needed:

gender1 <- cov1.dat[,2]

program1 <- cov1.dat[,3]

smoking1 <- cov1.dat[,4]

drugs1 <- cov1.dat[,5]

```


```{r,echo=FALSE,message=FALSE,warning=FALSE,results='hide'}

#descriptive analysis about the network 4&5 before splitting it into positive and negative networks

#network 4

#indegree mean & SD

network4_indeg_mean <- round(mean(degree(t4.data,cmode = "indegree")), digits=2)

network4_indeg_SD <- round(sd(degree(t4.data,cmode = "indegree")), digits=2)

#Outdegree mean & SD

network4_outdeg_mean <- round(mean(degree(t4.data,cmode = "outdegree")), digits=2)

network4_outdeg_SD <- round(sd(degree(t4.data,cmode = "outdegree")), digits=2)

#Tranasitivity

network4_post4  <- round(gtrans(t4.data, measure = "weak"),digits = 2)

#reciprocity

network4_recip <- round(grecip(network4,measure="edgewise"),digits = 2)

#density
density_network4 <- round(network.density(network4, na.omit=TRUE, discount.bipartite=FALSE),digits=2)

########################

#network 5

#indegree mean & SD

network5_indeg_mean <- round(mean(degree(t5.data,cmode = "indegree")), digits=2)

network5_indeg_SD <- round(sd(degree(t5.data,cmode = "indegree")), digits=2)

#Outdegree mean & SD

network5_outdeg_mean <- round(mean(degree(t5.data,cmode = "outdegree")), digits=2)

network5_outdeg_SD <- round(sd(degree(t5.data,cmode = "outdegree")), digits=2)

#Tranasitivity

network5_post4  <- round(gtrans(t5.data, measure = "weak"),digits = 2)

#reciprocity

network5_recip <- round(grecip(network5,measure="edgewise"),digits = 2)

#density
density_network5 <- round(network.density(network5, na.omit=TRUE, discount.bipartite=FALSE),digits=2)


#################

#put them all in a data frame

#network 4 

net44_df <- t(data.frame(
  Indegree_mean = network4_indeg_mean,
  Outdegree_mean = network4_outdeg_mean,
  Indegree_SD = network4_indeg_SD,
  Outdegree_SD = network4_outdeg_SD,
  Transitivity = network4_post4,
  Reciprocity = network4_recip,
  Density = density_network4 ))

#network 5 

net55_df <- t(data.frame(
  Indegree_mean = network5_indeg_mean,
  Outdegree_mean = network5_outdeg_mean,
  Indegree_SD = network5_indeg_SD,
  Outdegree_SD = network5_outdeg_SD,
  Transitivity = network5_post4,
  Reciprocity = network5_recip,
  Density = density_network5 ))

#merge them all


Merged4455_df <- cbind(net44_df,net55_df)

```



```{r,echo=FALSE,message=FALSE,warning=FALSE}

#ploting the networks development

par(mfrow=c(1,3))

plot(network1,xlab="Network 1")
plot(network4,xlab="Network 4")
plot(network5,xlab="Network 5")

mtext("Networks Development Over Time ", outer=TRUE,  cex=2, line=-3)

```


```{r, results='asis',message=FALSE,echo=FALSE,warning=FALSE}

#add them in kable

library(knitr)

kable(Merged4455_df ,
      col.names = c(" Network 4"," Network 5"),
      caption = "Descriptive network statistics for networks 4 and 5")
      

```




```{r,echo=FALSE,message=FALSE,warning=FALSE,results='hide'}
#descriptive analysis for positive and negative networks 4 and 5 

############
#Positive network 4 

#indegree mean & SD

post4_indeg_mean <- round(mean(degree(Positive_data4,cmode = "indegree")), digits=2)

post4_indeg_SD <- round(sd(degree(Positive_data4,cmode = "indegree")), digits=2)

#Outdegree mean & SD

post4_outdeg_mean <- round(mean(degree(Positive_data4,cmode = "outdegree")), digits=2)

post4_outdeg_SD <- round(sd(degree(Positive_data4,cmode = "outdegree")), digits=2)

#Tranasitivity

trans_post4  <- round(gtrans(Positive_data4, measure = "weak"),digits = 2)

#reciprocity

post4_recip <- round(grecip(network4_positive,measure="edgewise"),digits = 2)

#density
density_post4 <- round(network.density(network4_positive, na.omit=TRUE, discount.bipartite=FALSE),digits=2) 

############

#negative network 4 

#indegree mean & SD

negat4_indeg_mean <- round(mean(degree(negative_data4,cmode = "indegree")), digits=2)

negat4_indeg_SD <- round(sd(degree(negative_data4,cmode = "indegree")), digits=2)

#Outdegree mean & SD

negat4_outdeg_mean <- round(mean(degree(negative_data4,cmode = "outdegree")), digits=2)

negat4_outdeg_SD <- round(sd(degree(negative_data4,cmode = "outdegree")), digits=2)

#Tranasitivity

trans_negat4  <- round(gtrans(negative_data4, measure = "weak"),digits = 2)

#reciprocity

negat4_recip <- round(grecip(network4_negative,measure="edgewise"),digits = 2)

#density

density_negat4 <- round(network.density(network4_negative, na.omit=TRUE, discount.bipartite=FALSE),digits=2) 

#all in df
network4_df <- t(data.frame(
  Indegree_mean = c(post4_indeg_mean,negat4_indeg_mean),
  Outdegree_mean = c(post4_outdeg_mean,negat4_outdeg_mean),
  Indegree_SD = c(post4_indeg_SD,negat4_indeg_SD),
  Outdegree_SD = c(post4_outdeg_SD,negat4_outdeg_SD),
  Transitivity = c(trans_post4,trans_negat4),
  Reciprocity = c(post4_recip,negat4_recip),
  Density = c(density_post4,density_negat4)))


```


```{r,echo=FALSE,message=FALSE,warning=FALSE,results='hide'}

#Positive network 5 

#indegree mean & SD

post5_indeg_mean <- round(mean(degree(Positive_data5,cmode = "indegree")), digits=2)

post5_indeg_SD <- round(sd(degree(Positive_data5,cmode = "indegree")), digits=2)

#Outdegree mean & SD

post5_outdeg_mean <- round(mean(degree(Positive_data5,cmode = "outdegree")), digits=2)

post5_outdeg_SD <- round(sd(degree(Positive_data5,cmode = "outdegree")), digits=2)

#Tranasitivity

trans_post5  <- round(gtrans(Positive_data5, measure = "weak"),digits = 2)

#reciprocity

post5_recip <- round(grecip(network5_positive,measure="edgewise"),digits = 2)

#density
density_post5 <- round(network.density(network5_positive, na.omit=TRUE, discount.bipartite=FALSE),digits=2) 

############

#negative network 5 

#indegree mean & SD

negat5_indeg_mean <- round(mean(degree(negative_data5,cmode = "indegree")), digits=2)

negat5_indeg_SD <- round(sd(degree(negative_data5,cmode = "indegree")), digits=2)

#Outdegree mean & SD

negat5_outdeg_mean <- round(mean(degree(negative_data5,cmode = "outdegree")), digits=2)

negat5_outdeg_SD <- round(sd(degree(negative_data5,cmode = "outdegree")), digits=2)

#Tranasitivity

trans_negat5  <- round(gtrans(negative_data5, measure = "weak"),digits = 2)

#reciprocity

negat5_recip <- round(grecip(network5_negative,measure="edgewise"),digits = 2)

#density

density_negat5 <- round(network.density(network5_negative, na.omit=TRUE, discount.bipartite=FALSE),digits=2) 

#all in df
network5_df <- t(data.frame(
  Indegree_mean = c(post5_indeg_mean,negat5_indeg_mean),
  Outdegree_mean = c(post5_outdeg_mean,negat5_outdeg_mean),
  Indegree_SD = c(post5_indeg_SD,negat5_indeg_SD),
  Outdegree_SD = c(post5_outdeg_SD,negat5_outdeg_SD),
  Transitivity = c(trans_post5,trans_negat5),
  Reciprocity = c(post5_recip,negat5_recip),
  Density = c(density_post5,density_negat5)))
############

#all df merged 

Merged_df <- cbind(network4_df,network5_df)

```


```{r, results='asis',message=FALSE,echo=FALSE,warning=FALSE}

#add them in kable

library(knitr)

kable(Merged_df,
      col.names = c("Positive Network 4","Negative Network 4","Positive Network 5","Negative Network 5"),
      caption = "Descriptive network statistics for students data set Marijtje van Duijn")


```



```{r,echo=FALSE,message=FALSE,warning=FALSE}

#plot positive and negative network 4 

par(mfrow=c(1,2))

plot(network4_positive,xlab=" Positive Ties")
plot(network4_negative,xlab=" Negative Ties")

mtext("Network 4 ", outer=TRUE,  cex=2, line=-3)
#plot positive and negative network 5

par(mfrow=c(1,2))

plot(network5_positive,xlab=" Positive Ties")
plot(network5_negative,xlab="Negative Ties")

mtext("Network 5 ", outer=TRUE,  cex=2, line=-3)

```





```{r,echo=FALSE,message=FALSE,warning=FALSE,results='hide'}

#ERGM models for the fifth positive network   

library(latticeExtra)	
library(ergm)
library(network)

#prepare the network and Program similarties 

Positive_Network <- network5_positive

#nodes only exist in the network 5 with the program variable 

Program <- program1[c(1,2,4,9,10,11,15,16,17,18,21,28,29,30,31,33,36,38)]

#prepare the gender 

Positive_Network %v% "Program" <- Program

#then create the first model 

model1 <- Positive_Network~
	edges+	# overall tendency to create ties
	mutual+		# reciprocation
	gwesp(0.5,fixed=TRUE)	+# transitive closure
   nodematch('Program',diff=F) #how Program matching affect the network formation? 

results1 <- ergm(model1)

summary(results1)


```



```{r,echo=FALSE,message=FALSE,warning=FALSE,results='hide'}
#negative network ERGM 

Negative_network <- network5_negative

Smoke <- smoking1[c(1,2,4,9,10,11,15,16,17,18,21,28,29,30,31,33,36,38)]

Drugs <- drugs1[c(1,2,4,9,10,11,15,16,17,18,21,28,29,30,31,33,36,38)]

Negative_network  %v% "Smoke" <- Smoke
Negative_network  %v% "Drugs" <- Drugs

model2 <- Negative_network~
	edges+	# overall tendency to create ties
	mutual+		# reciprocation
	twopath+	# precondition for transitive closure
	gwesp(0.5,fixed=TRUE)+	# transitive closure
   absdiff("Smoke") + #how smoke affect the network? 
  absdiff("Drugs") #how Drugs affect the network? 

results2 <- ergm(model2)

summary(results2)




```


```{r,echo=FALSE,message=FALSE,warning=FALSE,results='asis'}

library(stargazer)

stargazer(results1,results2, header = FALSE, title = "Two ERGM Models for Positive and Negative Networks ", notes = "Data taken from data set Marijtje van Duijn.")

```








```{r,echo=FALSE,message=FALSE,warning=FALSE,results='hide'}
#simulation

#for positive

Sim_positive <- simulate(results1,nsim=100)

#for negative 

Sim_negative <- simulate(results2,nsim=100)

```




```{r,echo=FALSE,message=FALSE,warning=FALSE}
#then, calculate density for positive simulated networks 

par(mfrow=c(1,2))
hist(gden(Sim_positive), xlab = "The Positive Model's Density", main = "")

# Add vertical line where empirical data are:
lines(rep(gden(Positive_Network),2),c(0,30),col="red",lwd=5)

#then, calculate density for negative simulated networks 

hist(gden(Sim_negative), xlab = "The Negative Model's Density",main = "")

# Add vertical line where empirical data are:
lines(rep(gden(Negative_network),2),c(0,30),col="red",lwd=5)

mtext("Density Of Simulated Networks From Both Models", outer=TRUE,  cex=1, line=-1.5)


```


```{r,echo=FALSE,message=FALSE,warning=FALSE}
#then, calculate reciprocity for the positive simulated networks

par(mfrow=c(1,2) )

hist(grecip(Sim_positive, measure = "edgewise"), xlab = "The Positive Model's Reciprocity", 
     main = "")

# Add vertical line where empirical data are:
lines(rep(grecip(Positive_Network, measure = "edgewise"),2),c(0,30),col="red",lwd=5)

#then, calculate reciprocity for the negative simulated networks

hist(grecip(Sim_negative, measure = "edgewise"),  xlim = c(-0.2,0.2), xlab = "The Negative Model's Reciprocity", main = "")

# Add vertical line where empirical data are:
lines(rep(grecip(Negative_network, measure = "edgewise"),2),c(0,30),col="red",lwd=5)

mtext("Reciprocity Of Simulated Networks From Both Models", outer=TRUE,  cex=1, line=-1.5)


```



```{r,echo=FALSE,message=FALSE,warning=FALSE}

#then, calculate transitivity for the positive simulated networks

 Sim1.transitivities <- sapply(Sim_positive,function(er){
	sna::gtrans((er))
})

par(mfrow=c(1,2))

hist(Sim1.transitivities, xlab = "The Positive Model's Transitivity",xlim = c(0.3,0.9), main = "")

# Add vertical line where empirical data are:
lines(rep(gtrans(Positive_Network, measure = "weak"),2),c(0,40),col="red",lwd=5)

#then, calculate transitivity for the negative simulated networks

Sim2.transitivities <- sapply(Sim_negative,function(er){
	sna::gtrans((er))
})

hist(Sim2.transitivities, xlim = c(-0.3,0.3), xlab = "The Negative Model's Transitivity", main = "")

# Add vertical line where empirical data are:
lines(rep(gtrans(Negative_network, measure = "weak"),2),c(0,100),col="red",lwd=5)


mtext("Transitivity Of Simulated Networks From Both Models", outer=TRUE,  cex=1, line=-1.5)

```


6.   Conclusions and discussion: 

Ultimately, the descriptive results and the ERG models show that the positive ties have a high tendency for density, reciprocity, and transitivity, as these ties associated with homophily such as program study, reflecting that the social networks always seek the positive actors with similar attributes or values to communicate with (Hypothesis  One). Moreover, the negative ties show a weak density, transitivity, and zero reciprocity that shows how the actors avoid reciprocating and sending edges in the negative network because of a lack of cognition in many networks (Hypothesis  Two). Nonetheless, for future research, we can investigate how the negative ties developed from a positive to a negative tie and what attributes we can consider the effect this change, as it seems important to investigate the positive and negative ties under many conditions. 







*References: 

Harrigan, N. and Yap, J., 2017. Avoidance in negative ties: Inhibiting closure, reciprocity, and homophily. Social Networks, 48, pp.126-141. 

Huitsing, G., Van Duijn, M.A., Snijders, T.A., Wang, P., Sainio, M., Salmivalli, C. and Veenstra, R., 2012. Univariate and multivariate models of positive and negative networks: Liking, disliking, and bully–victim relationships. Social Networks, 34(4), pp.645-657.

Heider, F., 1946. Attitudes and cognitive organization. The Journal of psychology, 21(1), pp.107-112.
 
Lim, Y. and Rubineau, B., 2013. Social status and negative tie formation. In Academy of Management Proceedings (Vol. 2013, No. 1, p. 14355). Briarcliff Manor, NY 10510: Academy of Management.

Labianca, G. and Brass, D.J., 2006. Exploring the social ledger: Negative relationships and negative asymmetry in social networks in organizations. Academy of Management Review, 31(3), pp.596-614.

Lusher, D., Koskinen, J., & Robins, G. (Eds.). (2013). Exponential random graph models for social networks: Theory, methods, and applications. Cambridge University Press. Chapter 3 &5 

Newcomb, T.M., 1961. The acquaintance process as a prototype of human interaction.

Papachristos, A.V., Hureau, D.M. and Braga, A.A., 2013. The corner and the crew: The influence of geography and social networks on gang violence. American sociological review, 78(3), pp.417-447.

Veenstra, R., Steglich, C.E.G., 2012. Actor-based model for network and behavior dynamics: a tool to examine selection and influence processes. In: Laursen, B., Little, T.D., Card, N.A. (Eds.), Handbook of Developmental Research Methods. Guilford, New York, pp. 598–618.


Van de Bunt, G.G., M.A.J. van Duijn, and T.A.B. Snijders. 1999. Friendship networks through time: An actor-oriented statistical network model. Computational and Mathematical Organization Theory, 5, 167-192.

Van Duijn, Marijtje A.J., Zeggelink, Evelien P.H., Huisman, Mark, Stokman, Frans N., and Wasseur, Frans W. 2003. Evolution of sociology freshmen into a friendship network.
Journal of Mathematical Sociology 27, 153-191.



























